Weighted Model Counting Without Parameter Variables

### 2021-03-01 doc/paper3/paper.tex
Writing about experiments

\section{Experimental Evaluation}
% TODO: what are the encodings compiled with?

We run a set of experiments, comparing all five original Bayesian network
encodings (\texttt{bklm16}, \texttt{cd05}, \texttt{cd06}, \texttt{d02}
\texttt{sbk05}) as well as the first four with \cref{alg:transformation} applied
afterwards.\footnote{Recall that \texttt{cd05} and \texttt{cd06} are
  incompatible with \textsf{DPMC}.} For each encoding \texttt{e}, we will write
\texttt{e++} to denote the combination of encoding a Bayesian network as a WMC
instance using \texttt{e} and transforming it into a PBP instance using
\cref{alg:transformation}. Along with
\textsf{DPMC}\footnote{\url{https://github.com/vardigroup/DPMC}}, we also
include WMC algorithms used in the papers that introduce each encoding:
\textsf{Ace} for \texttt{cd05}, \texttt{cd06}, and \texttt{d02};
\textsf{Cachet}\footnote{\url{https://cs.rochester.edu/u/kautz/Cachet/}}
\cite{DBLP:conf/sat/SangBBKP04} for \texttt{sbk05}; and
\textsf{c2d}\footnote{\url{http://reasoning.cs.ucla.edu/c2d/}}
\cite{DBLP:conf/ecai/Darwiche04} with
\textsf{query-dnnf}\footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}}
for \texttt{bklm16}. We focus on the following questions:
\begin{itemize}
\item Can parameter variable elimination improve inference speed?
\item How does DPMC combined with encodings without (and with) parameter
  variables compare with other WMC algorithms and other encodings?
\item Which instances is our approach particularly successful on (compared to
  other algorithms and encodings and to the same encoding before our
  transformation)?
\item What proportion of variables is typically removed?
\item Do some encodings benefit from this transformation more than others?
\end{itemize}

\subsection{Setup (mostly done)}

@@

\paragraph{Notes.}
\begin{itemize}
\item \Cref{fig:cumulative} shows:
  \begin{itemize}
  \item $\textsf{DPMC}+\texttt{bklm16++}$ remains the best-performing
    combination across all time limits up to \SI{1000}{\second} with $\textsf{Ace} +
    \texttt{cd06}$ and $\textsf{DPMC}+\texttt{bklm16}$ not far behind;
  \item $\textsf{DPMC}+\texttt{bklm16++}$ to be 3.35 times faster than
    $\textsf{DPMC}+\texttt{bklm16}$ and 2.96 times faster than
    $\textsf{Ace}+\texttt{cd06}$.
  \end{itemize}
\item \Cref{tbl:performance} further shows that
  $\textsf{DPMC}+\texttt{bklm16++}$ solves almost a hundred more instances than
  any other combination, and is the fastest in \SI{69.1}{\percent} of them.
\item TODO: now talk about \cref{fig:scatter}.
\item We also record numbers of variables in each encoding before and after our
  additional transformation.
\item TODO: describe \cref{fig:box}.
\item Indeed, the numbers of variables in \texttt{bklm16++} and \texttt{d02++}
  encodings are equal for binary Bayesian networks (i.e., most of our data).
  Nonetheless, \texttt{bklm16++} performs much faster than \texttt{d02++}.
\end{itemize}
% TODO: Describe the results. Make sure to answer all of the listed questions.
% TODO: double check that the numbers in the table make sense.

@@

>>>>

\section{Experimental Evaluation (mostly done)}
% TODO: what are the encodings compiled with?

We run a set of experiments, comparing all five original Bayesian network
encodings (\texttt{bklm16}, \texttt{cd05}, \texttt{cd06}, \texttt{d02}
\texttt{sbk05}) as well as the first four with \cref{alg:transformation} applied
afterwards.\footnote{Recall that \texttt{cd05} and \texttt{cd06} are
  incompatible with \textsf{DPMC}.} For each encoding \texttt{e}, we will write
\texttt{e++} to denote the combination of encoding a Bayesian network as a WMC
instance using \texttt{e} and transforming it into a PBP instance using
\cref{alg:transformation}. Along with
\textsf{DPMC}\footnote{\url{https://github.com/vardigroup/DPMC}}, we also
include WMC algorithms used in the papers that introduce each encoding:
\textsf{Ace} for \texttt{cd05}, \texttt{cd06}, and \texttt{d02};
\textsf{Cachet}\footnote{\url{https://cs.rochester.edu/u/kautz/Cachet/}}
\cite{DBLP:conf/sat/SangBBKP04} for \texttt{sbk05}; and
\textsf{c2d}\footnote{\url{http://reasoning.cs.ucla.edu/c2d/}}
\cite{DBLP:conf/ecai/Darwiche04} with
\textsf{query-dnnf}\footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}}
for \texttt{bklm16}. We focus on the following questions:
\begin{itemize}
\item Can parameter variable elimination improve inference speed?
\item How does DPMC combined with encodings without (and with) parameter
  variables compare with other WMC algorithms and other encodings?
\item Which instances is our approach particularly successful on (compared to
  other algorithms and encodings and to the same encoding before our
  transformation)?
\item What proportion of variables is typically eliminated?
\item Do some encodings benefit from this transformation more than others?
\end{itemize}

\subsection{Setup}

@@

@@

\Cref{fig:cumulative} shows $\textsf{DPMC}+\texttt{bklm16++}$ to be the
best-performing combination across all time limits up to \SI{1000}{\second} with
$\textsf{Ace} + \texttt{cd06}$ and $\textsf{DPMC}+\texttt{bklm16}$ not far
behind. Overall, $\textsf{DPMC}+\texttt{bklm16++}$ is 3.35 times faster than
$\textsf{DPMC}+\texttt{bklm16}$ and 2.96 times faster than
$\textsf{Ace}+\texttt{cd06}$. \Cref{tbl:performance} further shows that
$\textsf{DPMC}+\texttt{bklm16++}$ solves almost a hundred more instances than
any other combination, and is the fastest in \SI{69.1}{\percent} of them.

The scatter plots in \cref{fig:scatter} show that how $\textsf{DPMC} +
\texttt{bklm16++}$ (and perhaps \textsf{DPMC} more generally) compares to
$\textsf{Ace} + \texttt{cd06}$ depends significantly on the data set: the former
is a clear winner on DQMR and Grid instances, while the latter performs well on
Mastermind and Random Blocks. Perhaps because the underlying WMC algorithm
remains the same, the difference between $\textsf{DPMC} + \texttt{bklm16}$ with
and without applying \cref{alg:transformation} is quite noisy, i.e, with most
instances scattered around the line of equality. However, our transformation
does enable \textsf{DPMC} to solve many instances that were previously beyond
its reach.

We also record numbers of variables in each encoding before and after applying
\cref{alg:transformation}. \Cref{fig:box} shows a significant reduction in the
number of variables. For instance, the median number of variables in instances
encoded with \texttt{bklm16} was reduced four times: from 1499 to 376. While
\texttt{bklm16++} results in the overall lowest number of variables, the
difference between \texttt{bklm16++} and \texttt{d02++} seems small. Indeed, the
numbers of variables in these two encodings are equal for binary Bayesian
networks (i.e., most of our data). Nonetheless, \texttt{bklm16++} is still much
faster than \texttt{d02++} when run with \textsf{DPMC}.

Overall, transforming WMC instances to the PBP format allows us to significantly
simplify each instance. This transformation is particularly effective on
\texttt{bklm16}, allowing it to surpass \texttt{cd06} and become the new state
of the art. While there is a similarly significant reduction in the number of
variables for \texttt{d02}, the performance of $\textsf{DPMC}+\texttt{d02}$ is
virtually unaffected. Finally, while our transformation makes it possible to use
\texttt{cd05} and \texttt{cd06} with \textsf{DPMC}, the two combinations remain
inefficient.
